Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-20 paper.tex
Pass through the Robustness section

% % % % % % % % % % % % % % % % % %
\subsection*{Illustrative examples}
% likelihood surface
% The last paragraph aims to highlight the flexibility of this approach 
% and the impact of the various
% degrees of completeness of the ARG on the likelihood.
% Note that the figures in this paragraph only serve this exact purpose,
% we are not claiming to have designed a new parameter estimation approach.

Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
collection of nodes (ancestral haplotypes that lived at some time 
$t$ in the past) and edges (a record that child node $c$ 
inherited the genomic interval $[x, y)$ from parent node $p$)
under the SMC. Other than some basic consistency constraints (e.g., parents
must be older than children) there are no requirements on the topology
described, and in particular, no requirement that the events from the 
model (the SMC) should be explicitly identified in the graph. Our 
formulation reasons about the events that \emph{could have resulted in}
the observed nodes and edges, and therefore frees us from the requirement
of having to explicitly write those down. This gives a great deal 
of flexibility, which we illustrate in this section via some examples.

The first example concerns the simplest possible polytomy (see
the Discussion for why we consider polytomies important).
Suppose we have a graph consisting of four nodes, $c_1,c_2,c_3,p$ and three edges
${(p, c_j, 0, 1): 1 \leq j \leq 3}$. The child nodes are samples at time
zero, and the parent has time $1$. The likelihood of this ARG is undefined
under HKY because the three-way star topology described cannot be interpreted as 
a sequence of binary coalescence events. However, it is clearly 
\emph{consistent with} such a sequence of events, which we can reason
about probabilistically. In our formulation we [Peter/Gertjan, help me 
out any hold my hand a bit here. How exactly to we assign a likelihood
to such a polytomy? Why do I get a larger likelihood for the ternary
tree described above than an example in which I break the polytomy? 
Naively, I would have thought that the likelihood should have been 
lower, but I clearly don't understand likelihoods...]
\comment{The likelihood of the tree described is $e^{-3rt_p-3\lambda t_p} \lambda^2$;
if the polytomy was broken by adding another node at $(1-u)t_p/2$ then
the likelihood would be $e^{-(3-u)rt_p - (3-2u)\lambda t_p} \lambda^2$.
The latter is larger because there's less total area in the tree
for recombination to have not happened and less time over which coalescence hasn't happened.
However, these are different trees, so I'm not sure what the point is?}

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/likelihood_surface}
    \caption{TODO add A/B and change ``unary regions''.
    Log likelihood-ratio curves for ARGs simulated under the SMC with
    various degrees of completeness. The likelihood ratio is computed with
    respect to the true recombination rate. In case of the full ARG the likelihood
    under the SMC has been computed using the exact time to the recombination event.
    Simulation performed with \texttt{msprime}, parameters:
    genome length 1Mb, 20 diploid samples,
    $r=2.5\times10^{-8}$, $N_e=1\times10^4$.}
    \label{fig:lik-surface}
\end{figure}

Our second example shows a different aspect of this robustness to missing 
data. In Fig~\ref{fig:lik-surface}A we can see the likelihood-ratio curves
obtained when we compute the likelihood of an example simulation,
in which the ARG is at three different levels of completeness
(illustrated in Fig~\ref{fig:lik-surface}B).
We first run a simulation under the SMC
using the \texttt{record\_full\_arg} option in \texttt{msprime}. This 
ARG explicitly identifies all of the recombination and common 
ancestor events that occurred in the simulation, and 
we can therefore compute a likelihood under the HKY formulation
(for example, using the \texttt{msprime.log\_arg\_likelihood} 
function). The example full ARG in 
Fig~\ref{fig:lik-surface}B has three nodes that are ``locally
unary''~\citep{wong_general_2023,fritze2024forest}. Nodes 6 and 7 correspond to the 
left and right parents of the recombination event causing the 
breakpoint at position 536~\cite[we need two nodes to uniquely identify
 the passage of ancestral material through recombination 
events; see][]{baumdicker_efficient_2021}, and therefore present
in only those trees. 
Node 8 is a coalescent node in the right-hand tree, but unary
in the left-hand tree, marking the passage of the lineage through
this ancestral haplotype (similar to node g in Fig~\ref{fig:smc-unary}).
Fig~\ref{fig:lik-surface}A shows that, remarkably, when we 
remove the recombination nodes (e.g., 6/7) but keep the locally unary nodes (e.g., 8)
and compute the likelihood using eq.~\eqref{eq:full-lik},
there is very little difference in the likelihood ratio
curve from the exact value with complete input. 
Most importantly, the minimum occurs at the true value of the 
recombination parameter.
When we further simplify
the ARG to remove all unary nodes~\citep{wong_general_2023,kelleher_efficient_2018} 
we obtain an ARG in the standard format used by \texttt{msprime}.
Although there is substantial information about the correlation
structure of trees in this representation, we lose important
information about recombination. This is demonstrated in 
Fig~\ref{fig:lik-surface}A, where the corresponding likelihood ratio curve
is clearly biased.

@@

All scripts and data used for this manuscript are available at
\url{https://github.com/gertjanbisschop/smc-bit-paper},
and an implementation of the algorithm is available at
\url{https://github.com/gertjanbisschop/runsmc}.
\FloatBarrier
\bibliographystyle{plainnat}
\bibliography{refs.bib,paper.bib}

>>>>

% % % % % % % % % % % % % % % % % %
\subsection*{Robustness to partial information}
Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
collection of nodes (ancestral haplotypes that lived at some time 
$t$ in the past) and edges (a record that child node $c$ 
inherited the genomic interval $[x, y)$ from parent node $p$)
under the SMC. Other than some basic consistency constraints (e.g., parents
must be older than children) there are no requirements on the topology
described, and in particular, no requirement that the events from the 
model (the SMC) should be explicitly identified in the graph. Our 
formulation reasons about the events that \emph{could have resulted in}
the observed nodes and edges, and therefore frees us from the requirement
of having to explicitly write those down. In this section we 
illustrate the robustness
of this approach to incomplete information about recombination and 
coalescent events via a proof-of-concept example, in which we 
remove various levels of detail from a simulated ARG.

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/likelihood_surface}
    \caption{Log likelihood-ratio curves for ARGs containing
     varying degrees of information about the realised simulation.
    The ``Full ARG'' here is the result of an 
    \texttt{msprime} simulation of the SMC (100 diploid samples, genome length 1Mb, 
    $r=2.5\times10^{-8}$, $N_e=1\times10^4$) which contains complete 
    information about the simulated events.
    with various degrees of completeness. The likelihood ratio is computed with
    respect to the true recombination rate. 
    \label{fig:lik-surface}}
\end{figure}

In Fig~\ref{fig:lik-surface} we begin with a realisation of the SMC as a ``full
ARG'', that is, containing complete information about every simulated event.
We then plot the likelihood computed by Eq.~\eqref{eq:full-lik} 
as we vary the recombination rate parameter $r$ (as a ratio
of the likelihood computed at the true value of $r$). Unsurprisingly, we can
see that the minimum of the likelihood-ratio curve when using the 
full simulated ARG is at the true value of the parameter.
What is surprising is that we obtain an indistinguishable curve when 
we remove all the recombination nodes from this ARG, and therefore 
lose information about the precise timing of recombination events.
In this case, the locally unary portions of coalescent nodes have given
all the information that we need about recombination events.
Removing all locally unary nodes in the fully simplified ARG does result
in a loss of information, and we can see that the minimum of the 
likelihood ratio surface is biased away from its true value.

% Our second example shows a different aspect of this robustness to missing 
% data. In Fig~\ref{fig:lik-surface}A we can see the likelihood-ratio curves
% obtained when we compute the likelihood of an example simulation,
% in which the ARG is at three different levels of completeness
% (illustrated in Fig~\ref{fig:lik-surface}B).
% We first run a simulation under the SMC
% using the \texttt{record\_full\_arg} option in \texttt{msprime}. This 
% ARG explicitly identifies all of the recombination and common 
% ancestor events that occurred in the simulation, and 
% we can therefore compute a likelihood under the HKY formulation
% (for example, using the \texttt{msprime.log\_arg\_likelihood} 
% function). The example full ARG in 
% Fig~\ref{fig:lik-surface}B has three nodes that are ``locally
% unary''~\citep{wong_general_2023,fritze2024forest}. Nodes 6 and 7 correspond to the 
% left and right parents of the recombination event causing the 
% breakpoint at position 536~\cite[we need two nodes to uniquely identify
%  the passage of ancestral material through recombination 
% events; see][]{baumdicker_efficient_2021}, and therefore present
% in only those trees. 
% Node 8 is a coalescent node in the right-hand tree, but unary
% in the left-hand tree, marking the passage of the lineage through
% this ancestral haplotype (similar to node g in Fig~\ref{fig:smc-unary}).
% Fig~\ref{fig:lik-surface}A shows that, remarkably, when we 
% remove the recombination nodes (e.g., 6/7) but keep the locally unary nodes (e.g., 8)
% and compute the likelihood using eq.~\eqref{eq:full-lik},
% there is very little difference in the likelihood ratio
% curve from the exact value with complete input. 
% Most importantly, the minimum occurs at the true value of the 
% recombination parameter.
% When we further simplify
% the ARG to remove all unary nodes~\citep{wong_general_2023,kelleher_efficient_2018} 
% we obtain an ARG in the standard format used by \texttt{msprime}.
% Although there is substantial information about the correlation
% structure of trees in this representation, we lose important
% information about recombination. This is demonstrated in 
% Fig~\ref{fig:lik-surface}A, where the corresponding likelihood ratio curve
% is clearly biased.

@@

All scripts and data used for this manuscript are available at
\url{https://github.com/gertjanbisschop/smc-bit-paper},
and an implementation of the algorithm is available at
\url{https://github.com/gertjanbisschop/runsmc}.
\FloatBarrier
\bibliographystyle{plainnat}
\bibliography{paper.bib}
